---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Unzip and load the data into an R dataset.
```{r echo=TRUE, results="hide"}
path = "C:\\Users\\Joe\\Git\\RepData"
unzip(paste(path,"activity.zip",sep="\\"), exdir = path)
steps = read.csv(paste(path,"activity.csv",sep="\\"), na.strings="NA", colClasses=c("numeric","character","numeric"))
```
Convert the character field into a date
```{r echo=TRUE, results="hide"}
steps$date1 = as.Date(steps$date, format="%Y-%m-%d")
```
Load libraries to be used 
```{r echo=TRUE, results="hide"}
library(dplyr)
library(ggplot2)
```


## What is mean total number of steps taken per day?
Compute the mean and median and show a histogram of the total number of steps per day.
```{r echo=TRUE, results="hide"}
totalSteps = steps %>% group_by(date1) %>% summarise(steps=sum(steps, na.rm=TRUE)) %>% arrange(date1)
hist(totalSteps$steps, main="Number of Steps per Day", xlab="Number of steps")
meanSteps = round(mean(totalSteps$steps),0)
medianSteps = as.integer(round(median(totalSteps$steps),0))
```
The mean number of steps per day is `r meanSteps`.  
The median number of steps per day is `r medianSteps`.


## What is the average daily activity pattern?
Calculate the average number of steps per 5 minute interval across all days.
```{r echo=TRUE} 
avgIntvlSteps = steps %>% group_by(interval) %>% summarise(steps=mean(steps, na.rm=TRUE)) %>% arrange(interval)
plot(avgIntvlSteps$interval, avgIntvlSteps$steps, type="l", xlab="Time Interval (min)", 
     ylab="Number of Steps", main="Average Number of Steps per Time Interval")
```

Find the maximum number of steps and the corresponding time interval.
```{r echo=TRUE}
maxNumSteps = round(max(avgIntvlSteps$steps),0)
maxTimeInt = avgIntvlSteps$interval[which.max(avgIntvlSteps$steps)]
```
The time interval `r maxTimeInt` contains the maximum number of steps (`r maxNumSteps` steps) (averaged across all days).


## Imputing missing values
Count the number of rows with NAs, and the total number of NAs in the dataset.
```{r echo=TRUE}
rowNas = sum(is.na(steps[,1]) | is.na(steps[,2]) | is.na(steps[,3]))
totalNas = sum(is.na(steps))
```
The number of rows with at least one NA value in the row is `r rowNas`.  
The total number of NAs in the data is `r totalNas`.    


Create a new dataset with missing values replaced with the average for that interval.
```{r echo=TRUE}
stepsNoNas = steps
stepsNoNas[is.na(steps$steps),]$steps = 
  sapply(steps[is.na(steps$steps),]$interval, function(x) round(avgIntvlSteps[avgIntvlSteps$interval == x,]$steps,0))
```
Show a histogram of the total number of steps per day.
```{r echo=TRUE}
totalStepsNoNas = stepsNoNas %>% group_by(date1) %>% summarise(steps=sum(steps, na.rm=TRUE)) %>% arrange(date1)
hist(totalStepsNoNas$steps)
```

Compute the mean and median of the number of steps per day.
```{r echo=TRUE}
meanStepsNoNas = as.integer(round(mean(totalStepsNoNas$steps),0))
medianStepsNoNas = as.integer(round(median(totalStepsNoNas$steps),0))
sumStepsNoNas = as.integer(sum(round(avgIntvlSteps$steps,0)))
```
The mean number of steps per day (with NAs replaced) is `r meanStepsNoNas`.  
The median number of steps per day (with NAs replaced) is `r medianStepsNoNas`.  
The imputation approach has caused the mean and median to increase, and has increased the number of days containing `r sumStepsNoNas` steps, as this is the sum of the average number of steps for each time interval. This has caused a spike in the histogram around that value.  

## Are there differences in activity patterns between weekdays and weekends?
Create a factor variable to show if the data is for a weekday or a weekend.
```{r echo=TRUE}
weekends = function(dates){ factor(c("Weekday","Weekend")[as.numeric(weekdays(dates) %in% c("Saturday","Sunday"))+1]) }
stepsNoNas$weekend = weekends(stepsNoNas$date1)
```
Summarise data by weekday and time interval, and plot.
```{r echo=TRUE}
avStepsNoNas = stepsNoNas %>% group_by(weekend, interval) %>% summarise(steps=mean(steps, na.rm=TRUE)) %>% arrange(weekend, interval)
qplot(data=avStepsNoNas, x=interval, y=steps, facets=weekend~., geom="line", xlab="Time Interval", ylab="Number of steps", main="Average Number of Steps per Time Interval")
```
